---
title: "Which child car seats does the NHTSA rate highly?"
author: "fboulnois"
date: '2022-06-26'
output: rmarkdown::github_document
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'README.md')) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(jsonlite)
library(curl)

# load json from nhtsa api
load_from_api <- function(url) {
  results <- list()
  while(!is.null(url)) {
    response <- curl_fetch_memory(url, handle = new_handle())
    if(response$status_code != 200) {
      stop("Error reading nhtsa api")
    }
    json <- fromJSON(readBin(response$content, character()))
    url <- json$meta$pagination$nextUrl
    results <- append(results, list(json$result))
  }
  rbindlist(results)
}

# load the file cache, otherwise download from api
load_from_cache <- function(file, url) {
  if(file.exists(file)) {
    df <- fromJSON(file)
  } else {
    df <- load_from_api(url)
    write(toJSON(df, na = "null", pretty = 2), file=file)
  }
  df
}

# recursively expand rows that have lists of data.frames
expand <- function(df) {
  dt <- as.data.table(df)
  if(any(vapply(dt, is.list, logical(1)))) {
    dt <- rbindlist(apply(dt, 1, expand), fill = TRUE)
  }
  dt
}

# add columns to simplify notes and to count and group modes
add_cols <- function(dt) {
  dt[, modes.ratings.notes := paste0(na.omit(modes.ratings.notes), collapse = "\n"), by = id]
  dt[, modes.count := length(unique(modes.mode)), by = id]
  dt[, grp := .GRP, by = .(id, modes.mode)]
}

# reorder columns so that they are id, grp, ..., modes.mode, modes.count, ...
reorder_cols <- function(dt) {
  cols <- colnames(dt)
  b1 <- cols == "id" | cols == "grp"
  b2 <- cols == "modes.mode"
  b3 <- cols == "modes.count"
  b4 <- seq_along(cols) < which(b2) & !b1 & !b3
  b5 <- seq_along(cols) > which(b2) & !b1 & !b3
  newcols <- c(cols[b1], cols[b4], cols[b2], cols[b3], cols[b5])
  setcolorder(dt, newcols)
}

# add and reorder columns
modify_cols <- function(dt) {
  reorder_cols(add_cols(dt))
}
```

## Summary

Many infant and child car seats are available on Amazon, but which are truly
safe? The [National Highway Traffic Safety Administration](https://www.nhtsa.gov)
has a [Car Seat Finder tool](https://www.nhtsa.gov/equipment/car-seats-and-booster-seats)
but unfortunately it is difficult to compare and contrast car seats across
brands, features, and ratings.

To investigate in more detail I downloaded the raw NHTSA car seat data and
analyzed it using R. I was also interested in highly rated car seats that are
part of travel systems. These travel systems usually have a stroller and infant
car seat combo but these seats are almost never convertible / multifunction. I
cross-referenced the seat model of the most popular travel systems with the
NHTSA data.

The car seat data was analyzed in a few different dimensions:

-   Which car seats have the highest combined rating by geometric score
-   Which convertible (multifunction) car seats have the highest score
-   Which convertible car seats have an overall rating of 5 in any category
-   Which travel system car seats have the highest score

A few organizations also perform independent testing of car seats that meet or
exceed the NHTSA standards:

-   [BabyGearLab](https://www.babygearlab.com/t/car-seats)
-   [Consumer Reports](https://www.consumerreports.org/cro/car-seats.htm)

## Methods

### Data preparation

To keep this section short, the code which downloads and cleans the data has
been hidden, however it can be viewed as part of the raw `.Rmd` file that
produced this analysis.

Concisely, the NHTSA has a child seat [API](https://en.wikipedia.org/wiki/API)
which returns [JSON](https://en.wikipedia.org/wiki/JSON). Because that JSON has
a highly nested structure it is expanded into separate rows. To help with the
analysis, a few columns of data are added and some columns are reordered.

To avoid having to download the data every time, the data is loaded from a file
if it exists otherwise it downloads the data from the API.

```{r prepare}
url <- "https://api.nhtsa.gov/childSeats?offset=0&max=100&sort=make&order=asc&data=modes&dataSet=ratings"
json <- "car-seat-results.json"

raw <- modify_cols(expand(load_from_cache(json, url)))
```

### Data analysis

The NHTSA rates car seats across several categories:

```{r rating-type}
raw[, unique(modes.ratings.type)]
```

For each rating category, the seats are given numerical values from 1 to 5
where 5 corresponds to 5 stars (i.e. the highest / best rating):

```{r rating-num}
raw[, range(modes.ratings.rating, na.rm = TRUE)]
```

Powerful transformations are easy to do once the data is in a `data.table`
format. Since each rating category is different and could have different ranges,
the [geometric mean](https://en.wikipedia.org/wiki/Geometric_mean) is used to
calculate a single global average score for each car seat instead of the
arithmetic mean. The `OverallRating` column is also used in this calculation as
it does not seem to be an average of the other ratings but is instead the NHTSA
reviewer's overall rating of the car seat. Next, the table is converted from
long to wide format to pull out the ratings into their own columns, the rows are
ordered by highest geometric score first, and a final csv file is created.

```{r geom-score}
raw[, geom.score := exp(mean(log(modes.ratings.rating))), by = id]

dt <- dcast(raw, id + modes.mode + ... ~ modes.ratings.type, value.var = "modes.ratings.rating")
setorder(dt, -geom.score, na.last = TRUE)

csv <- "car-seat-results.csv"
fwrite(dt, csv)
```

## Results

### Distribution of scores

The ratings are mainly clustered around a score of
`r dt[, round(median(geom.score, na.rm = TRUE), 2)]`, with some below average
ratings and only a few high ratings.

```{r score-distribution, echo=FALSE, results='asis'}
hist(dt[, geom.score], breaks = 15, main = "Distribution of geometric scores", xlab = "Score")
```

### Highest scoring car seats

The highest rated car seats in all categories by geometric score are as follows:

```{r highest-scoring-seats, echo=1, results='asis'}
res <- dt[geom.score > 4.5, .(make, productModel, seatType, geom.score)]
knitr::kable(res)
```

The brands GB, Doona, Cybex, Britax, and Graco top the list of the highest rated
car seats.

### Highest scoring convertible car seats

The highest rated convertible car seats by geometric scores are as follows:

```{r highest-scoring-multi-seats, echo=1, results='asis'}
res <- unique(dt[modes.count == 3 & geom.score > 3.5, .(make, productModel, seatType, geom.score)])
knitr::kable(res)
```

Some of the most popular convertible car seats appear in this list, including
the 4Ever DLX and Grows4Me by Graco, the One4Life by Britax, the City View by
Baby Jogger, and the EveryStage by EvenFlo.

### Highest 5-star rated convertible car seats

The highest rated convertible car seats with an overall rating of 5 in at least
one category are as follows:

```{r highest-scoring-5-star-multi-seats, echo=1, results='asis'}
res <- dt[modes.count == 3 & OverallRating == 5, .(make, productModel, seatType, geom.score)]
knitr::kable(res)
```

The Graco 4Ever DLX, Britax One4Life Clicktight, and Evenflo EveryFit top the
list for highest 5-star rated convertible car seats.

### Highest scoring travel system car seats

I manually compiled a list of the most popular travel systems and their car
seats from a couple of review sites and Amazon:

```{r highest-scoring-travel-system-seats, echo=FALSE, results='asis'}
tt <- data.table(
  brand = c("Chicco", "Evenflo", "Nuna", "Graco", "Baby Trend", "Britax", "UPPAbaby", "Baby Jogger", "Safety 1st", "Evenflo", "Graco", "Doona", "Century", "Maxi-Cosi"),
  travel.system = c("Bravo Trio", "Pivot", "TAVO", "Modes", "Expedition Jogger", "B-Free & B-Safe Gen2", "MESA", "City Mini GT2", "Smooth Ride", "Urbini Omni Plus", "FastAction Fold Sport Click Connect", "Car Seat & Stroller", "Stroll On", "Zelia Max 5-in-1"),
  seat.name = c("Keyfit 30 Magic", "SafeMax Infant Seat", "Pipa Lite LX", "Snugride Snuglock 35", "EZ Flex-Loc", "B-Safe Gen 2", "MESA", "City Go 2", "onBoard 35 Air 360", "LiteMax 35", "Snugride 35 Click Connect", "Infant Seat", "Carry On 35", "Mico Max Plus")
)
knitr::kable(tt)
```

Next this table is combined with the NHTSA data to get a final list of the
highest scoring travel system car seats:

```{r highest-scoring-travel-system-seats-combined, echo=1:3, results='asis'}
res <- merge(dt, tt, by.x = c("make", "productModel"), by.y = c("brand", "seat.name"))
res <- res[, .(make, travel.system, productModel, seatType, geom.score)]
setorder(res, -geom.score, na.last = TRUE)
knitr::kable(res)
```

Unsurprisingly, all of these travel systems only have infant car seats. The
three highest scoring travel systems are the Britax B-Free & B-Safe Gen 2 travel
system, the Doona Car Seat & Stroller, and the Baby Jogger City Mini GT2 with
the City Go 2 car seat.

## Conclusion

Many infant and child seats are available for purchase, but only a few are
highly rated by the NHTSA individually, as a convertible car seat, and as part
of a travel system.

All of the data used in the analysis is available in this repository. The raw
data is stored in [``r json``](`r json`), and the final data is stored in
[``r csv``](`r csv`).

Hopefully this analysis helps your family make the right decision!
